Nice guys or potential cooperators when keeping promises: an experimental and Bayesian account for two explanations

### 2019-07-29 article_english.Rmd
Creating latex template

---
title: "Nice guys or potential cooperators when keeping promises: an experimental and Bayesian account for two explanations"
author: "Said Jiménez"
date: "July 2019"
output: 
  pdf_document:
    number_sections: true
fontsize: 12pt
bibliography: promises.bib
csl: apa.csl
---

@@

One of the forms of communication that have received the most attention in psychological game theory is the promises, in an interaction between the sender and the recipient, it is believed that promises influence beliefs of the recipient, generating trust and cooperation [@Charness2006; @Vanberg2008]. However, there are situations where trust is betrayed, promises are broken or people deceive, for example, approximately 11% of people surveyed by the General Social Survey (GSS) in 2018 responded that they have had sex with someone different to their partner while they are married [@GSS2018]. Another example, according to the Washington Post, President Trump has said 10,796 false or misleading statements in 869 days in office [@Kessler2019].

@@

3. Finally, we will obtain two subsamples according to the cooperation rate of the subjects as was done in another study [@Baumgartner2009]. However, what in the aforementioned study was classified as a group of _honest_ and _dishonest_, we will show that in our sample it is not supported because if the groups differ in their payment rate it will be due to the difference in their level of commitment expressed in the promises.

@@

The structure described is repeated in 24 trials, however, in 4 of the trials the trustee can send a promise to the investor, the promises were that _always_, _mostly_, _sometimes_,  or _never_ would pay back. Each promise was valid for three trials so that 12 of the trials have the effect of the promise and the other 12 do not. It is worth mentioning that the promises are not made to a particular partner, but the subject expresses their level of commitment to pay back regardless of who their partners are in the next three trials in which the promise is valid.

@@

In the **anticipation/assignment phase**, subjects were told who their partner was for that trial (computer, stranger or friend) and was given the message that their partner was making their decision. Later in the **decision phase**, the subject was informed if his partner invested his $2 or not, he was reminded of his promise level (if they were trails with promises) and, in case of his partner had invested, he was asked to decide whether to pay back or not. Finally, in **feedback phase** payments for that trial were shown and the sequence was repeated.

@@

We analyzed 810 decisions about paying back or not (18 per subject), as well as 180 promises (4 per subject), 69% of the decisions,  were pay back, while the proportion of the promises selected were 5.4%, 16.2% 43.8%, 40.2%, respectively for promises _never_, _sometimes_, _mostly_ and _always_. Figure 2 shows the payment proportion depending on the levels of promises and partners.

Breaking promises, that is, that the subject had promised _always_ and subsequently decides not to pay back occurred only 12.4% of trials (19 times they did not pay back of 153 when promised _always_), of which, 8.50% (13/153) were trials with the computer, 3.92% (6/153) with the stranger and 0% (0/153) with the friend. It is essential to note that breaking the promise was not the most frequent decision, in 91.5% of trials when they promised _always_ and had the computer as a partner they decided to pay back.

@@

Figure 3 shows the posterior probability of breaking the promise depending on the partner, circles correspond to the medians of the distributions of the posterior estimates effects, the thick bar, and the thin bar correspond to the interval of 50% and 95% posterior probability. It is clear that the probability of breaking the promise decreases as social closeness increases, however, we calculate the reasons for evidence for the following hypotheses:

@@

On the other hand, Figure 6 shows the subsequent predictive distribution, compared to the payment rates of all individuals to the different partners and their variation by the level of promise. With the exception of the promise _never_, a monotonic positive effect of partners is observed at all levels of promise, however, it can be observed how the effect of social closeness varies depending on the strength of the promise, mainly for the decisions towards the computer.

@@

Our experiment is probably the first to include a partner without social closeness (_computer_), our data give evidence that the mere fact of considering that you play with a human diminishes the probability of breaking the promise. It is approximately 7.5 times more likely to break the promise to the computer than to the stranger, even though that partner was not known. In our study, zero social closeness is important because it allows us to explore the _intrinsic_ motivation to keep the promises, if humans keep their word because it is morally correct, we would anticipate that probability of breaking the promise was very low in the three partners and, particularly, the inference regarding the probability of breaking the promise to the computer would include zero.

@@

On the other hand, high social closeness allows us to explore the _instrumental_ motivation to keep promises. In our study, it is almost 9 times more likely to break the promise to the stranger than to the friend; If human beings keep their word to facilitate cooperation in the future, we would anticipate that with a partner of high social closeness (with whom they are very likely to cooperate, even after the experiment) there would be less chance of breaking the promise. In fact, in our study, the promise to friends was not broken on any occasion, so the inference, in this case, is obvious.

Our findings could be explained by the social norm called _conditional cooperation_, which indicates that the belief that other people cooperate at high levels also induces high levels of cooperation [@Fehr2018]. Thus, the lack of certainty regarding the decisions that the computer would make could explain the probability of breaking the promise that exists towards this partner. Similarly, the information that the subjects have regarding the behavior of the friend -even before the experiment- can explain the high levels of cooperation and keeping of promises towards this partner.

@@

To try to homogenize social closeness between our subjects and their partners, participants performed the task with a same-sex match friend considered close by themselves, in none case was it a relative and, although we understand that there may be same-sex couples and that the subjects may have hidden their relationship, they were emphasized that to be part of the study their companion could not be a romantic or sexual partner. Besides, Bayesian analysis of the IOS scale showed evidence in favor that the degree of social closeness between 30 of our 45 subjects and their friends was practically the same. Although the results of the two studies mentioned are practically in the same direction as the findings of our experiment, establishing the similarity in effects magnitudes is not possible due to the design differences, the performed tasks, and the statistical procedures used.

@@

On the other hand, the *Cumulative Models* allowed us to model the strength of the commitment expressed through promises. A very frequent problem that has been pointed out recently is that analyzing ordinal data with methods that assume that observations are _metric_ can lead to serious inference errors [@Liddell2018]. The *Cumulative Models* assume that the observed responses come from the categorization of a continuous latent variable [@Burkner2019], so in our study, we could identify that the groups with different payment proportions, which in other articles have been called _dishonest_ and _honest_, differ in the commitment they express through their promises. Considering that the tools to perform the *Cumulative Models* are relatively recent, it would be valuable to replicate published studies [e.g., @Baumgartner2009; @Baumgartner2013] considering the new alternatives of inference.

Although the *Hierarchical* and *Cumulative* models are not tools exclusively for Bayesian inference, their application from this approach represents several advantages compared to frequentist statistics. Classic problems of the previous one such as multiple comparisons, the decision of when to stop collecting subjects (stopping rule), or use of planned comparisons versus _post hoc_, are not factors that affect the support provided by the data to the hypothesis from the Bayesian approach [@Dienes2011]. In Bayesian inference, hypothesis testing makes formal use of probability to express the plausibility of theories, in our case we were able to obtain evidence ratios regarding the extent to which our data support our hypotheses, regardless of whether these were proposed _a priori_ or _post hoc_.

A limitation of this paper is the use of hypothetical rewards compared to real rewards. A relatively recent study on decision-making reported that there is more loss aversion when subjects have real monetary rewards compared to hypothetical in a risk task [@Xu2016]. However, other works do not report differences between the use of both types of rewards in self-control, temporal and social discount tasks [@Locey2011; @Johnson2006; @Rachlin2006]. Likewise, it can be argued that the results have theoretical congruence [@Fehr2018] and are in the same direction as other works that use real money [@Croson1996; @Glaeser2000], so there are not many reasons to expect that other types of rewards would modify our results.

>>>>

---
title: "Nice guys or potential cooperators when keeping promises: an experimental and Bayesian account for two explanations"
author: "Said Jiménez"
date: "July 2019"
output: 
  pdf_document:
    number_sections: true
    keep_tex: true
fontsize: 12pt
bibliography: promises.bib
csl: apa.csl
---

@@

One of the forms of communication that have received the most attention in psychological game theory is the promise, in an interaction between sender and recipient, it is believed that promises influence beliefs of the recipient, generating trust and cooperation [@Charness2006; @Vanberg2008]. However, there are situations where trust is betrayed, promises are broken or people deceive, for example, approximately 11% of people surveyed by the General Social Survey (GSS) in 2018 responded that they have had sex with someone different to their partner while they are married [@GSS2018]. Another example, according to the Washington Post, President Trump has said 10,796 false or misleading statements in 869 days in office [@Kessler2019].

@@

3. Finally, we will obtain two subsamples according to the cooperation rate of the subjects as was done in another study [@Baumgartner2009]. However, what in the aforementioned study was classified as a group of _honest_ and _dishonest_, we will show that in our sample it is not supported because if the groups differ in their payment rate it will be due to the difference in their level of commitment expressed by promises.

@@

The structure described is repeated in 24 trials, however, in 4 of the trials the trustee can send a promise to the investor, the promises were that _always_, _mostly_, _sometimes_,  or _never_ would pay back. Each promise was valid for three trials so that 12 of the trials have the effect of the promise and the other 12 do not. It is worth mentioning that the promises are not made to a particular partner, but the subject expresses their commitment level to pay back regardless of who their partners are in the next three trials in which the promise is valid.

@@

In the **anticipation/assignation phase**, subjects were told who their partner was for that trial (computer, stranger or friend) and was given the message that their partner was making their decision. Later in the **decision phase**, the subject was informed if his partner invested his $2 or not, he was reminded of his promise level (if they were trails with promises) and, in case of his partner had invested, he was asked to decide whether to pay back or not. Finally, in **feedback phase** payments for that trial were shown and the sequence was repeated.

@@

We analyzed 810 decisions about paying back or not (18 per subject), as well as 180 promises (4 per subject), 69% of the decisions,  were pay back, while the proportion of the promises selected were 5.4%, 16.2%, 43.8% and 40.2%, respectively for promises _never_, _sometimes_, _mostly_ and _always_. Figure 2 shows the payment proportion depending on the levels of promises and partners.

Breaking promises, that is, subject had promised _always_ and subsequently decides not to pay back occurred only 12.4% of trials (19 times they did not pay back of 153 when promised _always_), of which, 8.50% (13/153) were trials with the computer, 3.92% (6/153) with the stranger and 0% (0/153) with the friend. It is essential to note that breaking the promise was not the most frequent decision, in 91.5% of trials when they promised _always_ and had the computer as a partner they decided to pay back.

@@

Figure 3 shows the posterior probability of breaking the promise depending on the partner, circles correspond to the medians of the distributions of the posterior estimates effects, the thick bar, and the thin bar correspond to the interval of 50% and 95% posterior probability, respectively. It is clear that the probability of breaking the promise decreases as social closeness increases, however, we calculate the reasons for evidence for the following hypotheses:

@@

On the other hand, Figure 6 shows the posterior predictive distribution, compared to the payment rates of all individuals to the different partners and their variation by the level of promise. With the exception of the promise _never_, a monotonic positive effect of partners is observed at all levels of promise, however, it can be observed how the effect of social closeness varies depending on the strength of the promise, mainly for the decisions towards the computer.

@@

Our experiment is probably the first to include a partner without social closeness (_computer_), our data give evidence that the mere fact of considering that subjects play with a human diminishes the probability of breaking the promise. It is approximately 7.5 times more likely to break the promise to the computer than to the stranger, even though that partner was not known. In our study, zero social closeness is important because it allows us to explore the _intrinsic_ motivation to keep the promises, if humans keep their word because it is morally correct, we would anticipate that probability of breaking the promise was very low in the three partners and, particularly, the inference regarding the probability of breaking the promise to the computer would include zero.

@@

On the other hand, high social closeness allows us to explore the _instrumental_ motivation to keep promises. In our study, it is almost 9 times more likely to break the promise to the stranger than to the friend; if human beings keep their word to facilitate cooperation in the future, we would anticipate that with a partner of high social closeness (with whom they are very likely to cooperate, even after the experiment) there would be less chance of breaking the promise. In fact, in our study, the promise to friends was not broken on any occasion, so the inference, in this case, is obvious.

Our findings could be explained by the social norm called _conditional cooperation_, which indicates that the belief that other people cooperate at high levels also induces high levels of cooperation [@Fehr2018]. Thus, the lack of certainty regarding the decisions that the computer would make could explain the probability of breaking the promise that exists towards this partner. Similarly, the information that the subjects have regarding friend's behavior -even before the experiment- can explain the high levels of cooperation and keeping of promises towards this partner.

@@

To try to homogenize social closeness between our subjects and their partners, participants performed the task with a same-sex match friend considered close by themselves, in none case was it a relative and, although we understand that there may be same-sex couples and that the subjects may have hidden their relationship, they were emphasized that to be part of the study their companion could not be a romantic or sexual partner. Besides, Bayesian analysis of the IOS scale showed evidence in favor that the degree of social closeness between 30 of our 45 subjects and their friends was the same. Although the results of the two studies mentioned are practically in the same direction as the findings of our experiment, establishing the similarity in effects magnitudes is not possible due to the design differences, the performed tasks, and the statistical procedures used.

@@

On the other hand, the *Cumulative Model* allowed us to capture the strength of the commitment expressed through promises. A very frequent problem that has been pointed out recently is that analyzing ordinal data with methods that assume that observations are _metric_ can lead to serious inference errors [@Liddell2018]. The *Cumulative Models* assume that the observed responses come from the categorization of a continuous latent variable [@Burkner2019], so in our study, we could identify that the groups with different payment proportions, which in other articles have been called _dishonest_ and _honest_, differ in the commitment they express through their promises. Considering that the tools to perform the *Cumulative Models* are relatively recent, it would be valuable to replicate published studies [e.g., @Baumgartner2009; @Baumgartner2013] considering the new alternatives of inference.

Although the *Hierarchical* and *Cumulative* models are not tools exclusively for Bayesian inference, their application from this approach represents several advantages compared to frequentist statistics. Classic problems of the previous one such as multiple comparisons, the decision of when to stop collecting subjects (stopping rule), or use of planned comparisons versus _post hoc_, are not factors that affect the support provided by the data to the hypothesis from the Bayesian approach [@Dienes2011]. In Bayesian inference, hypothesis testing makes formal use of probability to express the plausibility of theories, in our case we were able to obtain evidence ratios regarding the extent our data support hypotheses, regardless these were proposed _a priori_ or _post hoc_.

A limitation of this paper is the use of hypothetical rewards compared to real rewards. A relatively recent study on decision-making reported that there is more loss aversion when subjects have real monetary rewards compared to hypothetical in a risk task [@Xu2016]. However, other works do not report differences between the use of both types of rewards in self-control, temporal and social discount tasks [@Locey2011; @Johnson2006; @Rachlin2006]. Likewise, it can be argued that our results have theoretical congruence [@Fehr2018] and are in the same direction as other works that use real money [@Croson1996; @Glaeser2000], so there are not many reasons to expect that other types of rewards would modify our results.
